Learning how to act: making good decisions with machine learning

### 2017-07-06 thesis/thesis_map.tex
mostly complets contextual bandits section

\chapter*{Todo}
\begin{enumerate}
\item Add something about Granger causality - does this belong in chapter 2 or chapter 3?
\item Sell my PhD as two key contribution. 1) A technical one, the other educational and unifying (accross viewpoints)
\item Add a chapter in 'PhD prose' making some statement about the intersection of prediction and causality. When is a problem causal, what bounds does predictive accuracy give on causal models? To what degree is predictive accuracy a metric on which causal models can be judged? 
\item Benchmark estimation algorithms
\item clarify causal interpretation of dirichlet models
\item Empirical evaluation of bandit algorithms (updated to include latest suggestions). Maybe a proper software benchmarking solution?
\item If comparing bandit performance, it would be interesting to include some that violate the stochastic bandit assumption. For example, if the rewards change slowly over time or jump suddenly a limited number of times, or with some small probability. 
\end{enumerate}

@@

This implies the counterfactual distributions are equivelent to the corresponding conditional distributions and, for a binary outcome $Y$, the causal effect is,

@@

The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignorebility assumptions \citep{Rosenbaum1983}. They state that the treatment each person receives is independent of whether they would recover if treated and if they would recover if not treated. This is justified in example \ref{exm:ranomized_experiment} due to the randomisation of treatment assignment. In general the treatment assignment will not be independent of the potential outcomes. In example \ref{exm:adjusting}, the children who attended preschool may be more likely to have graduated high school had they in fact not attended than the children who actually did not attend, $X \nci \cf{Y}{0}$. Similarly, had the poorer children who did not attend pre-school attended they might not have done as well as the children who did in fact attend, $X \nci \cf{Y}{1}$. A more general form of the ignorebility assumption is to identify a set of variables $Z$ such that $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$.

@@

\begin{figure}[h]
\centering
\caption{
Parametric regression may yeild poor estimates of causal effects if the model is miss-specified, even if the model fits well over the domain of the training data. In this example, $\P{Z|X=0} \sim N(\mu_0,\sigma_0^2)$ and $\P{Z|X=1} \sim N(\mu_1,\sigma_1^2)$ with little overlap in the densities. If $X=0$ then $Y \sim N(f_1(x)=sin(x),\sigma_y^2)$ and if $X=1$ then $Y \sim N(f_0(x) = \frac{1}{x+1},\sigma_y^2)$. We estimate $f_1(z)$ from the sample in which $X=1$ (green points) and $f_0(z)$ from the sample for which $X=0$ (blue points). In both cases the linear model is a good fit to the data. However, the resulting estimate of the causal effect is very poor for the lower values of $z$. } 
\label{fig:covariate_shift_adjust}
\includegraphics[scale=.55]{covariate_shift_estimation.pdf}
\end{figure}

@@

The propensity score can be used to estimate the average causal effect without specifying a regression model for $\E{Y|X,\vb{Z}}$. \citet{Rosenbaum1983} demonstrated that if the ignorability assumption is satisfied by conditioning on $\vb{Z}$, then it is also satisfied by conditioning on $e(\vb{z})$. This allows for estimators based on stratifying, matching or regression on the propensity score rather than the covariates $\vb{Z}$. Inverse propensity weighting can also be combined with empirical estimation of $\E{Y|X,Z}$ yielding the simple, albeit inefficient, estimator in equation \ref{eqn:simple_inverse_propensity} \citep{Imbens2004}. In some settings, such as stratified randomised trials \citep{imbens2015causal} or learning from logged bandit feedback \citep{Bottou2013} the propensity score may be known. However in general, it must be estimated from data. Frequently this is done with a simple parametric model such as logistic regression, but a wide range of standard machine learning algorithms including bagging and boosting, random forests and nueral networks can also be applied   \citep{Austin2011}. \citet{Lunceford2004} review the theoretical properties of key propensity score based estimators, including stratification and inverse propensity weighting.

@@

The performance of methods for causal effect estimation can be tested on simulated data \citep{Frolich2001,zhao2004using,Hill2011,Dorie2016} or by comparing estimates from observational studies with the results from corresponding exeriments \citep{lalonde1986evaluating,Fraker1987,heckman1997matching,heckman1998characterizing,dehejia1999causal,Smith2001,Anglemyer2014}. Unfortunately there are a relatively small number of examples where comparable observational and experimental data are available. The results are mixed with later studies finding generally better alignment of results but it is hard to ascertain if this is due to improved methodological approaches or over-fitting to the available data-sets.

@@

The previous sections all focus on aspects of the question; how can we estimate the effect of an intervention in a system from data collected prior to taking it. There is an obvious alternative. Instead of trying to infer the outcome of an intervention from passive observations one can intervene and see what happens. There are three key differences between observing a system and explicitly intervening in it. Firstly, when we intervene we choose which actions to take and thus have control over which data points we obtain to learn from. Selecting data points optimally for learning is the focus of the optimal experimental design liturature within statistics \citep{pukelsheim2006optimal} and the active learning liturature in machine learning \citep{settles2010active}. Secondly, explicitly choosing interventions yields a perfect model of the probability with which each action is selected, given any context, allowing control over confounding bias. Finally, when we are intervening in a system we typically care about the impact of our actions on the system in addition to optimising learning. For example, in a drug trial , assigning people a sub-optimal treatment has real world costs. This leads to a trade-off between exploiting the best known action so far and exploring alternative actions about which we are less certain. This exploration-exploitation trade-off lies at the heart of the field of reinforcement learning \citep{sutton1998reinforcement}.

@@

Stratified randomised experiments address the issue of variance due to covariate imbalance by randomly allocating treatment conditional on covariates belived to influence the outcome of interest. This introduces a backdoor path from treatment to outcome via the covariates on which treatment is stratified, see figure \ref{fig:random_experiment_network_stratified}, nessessitating that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under under ignorability \S\ref{sec:causal_effect_under_ignorability}. The key difference is that the propensity score is known, as it is designed by the experimenter, and there are guaranteed (rather than assumed) to be no latent confounding variables (that influence both treatment and outcome). See \citet{imbens2015causal} for a discussion of the trade-offs between statified versus completely random experiments.

@@

It is not always possible or ethical to conduct a randomized controlled trial as is beautifully demonstrated by the paper of \citet{smith2003parachute} on randomised cross-over trials of parashute use for the reduction of the mortality and morbity associated with falls from large hights \ref{fig:parashute}. When experimentation is possible, it is frequently difficult or expensive. This means experimental data sets are often much smaller than observational ones, limiting the complexity of models we can explore. In addition, they are often conducted on a convenient but unrepresentative sample of the broader population of interest (for example first year students at research universities). This can result in estimates with high \emph{internal validity} \citep{Campbell1963} in that they should replicate well in a similar population, but very low \emph{external validity}; the results may not carry over to the general population of interest. The question of whether an experiment conducted on one population can be mapped to another is referred to as the transportability problem \citep{Bareinboim2013} and relies on very similar assumptions and arguments to causal inference and the do-calculus.

\begin{figure}
\begin{center}
\caption{Experiments are not always ethical; an illustration of a randomised cross-over trial of parashutes for the prevention of morbitity and mortality associated with falls from large heights.}
\label{fig:parashute}
\includegraphics[scale=.4]{double_blind_parashute_trial_cartoon.png}
\end{center}
\end{figure}

@@

Multi-armed bandits address the problem of desiging experiments that can adapt as samples are observed. Their introduction is generally attributed to \citet{Thompson1993}. In its classic formulation \cite{Robbins1952,Lai1985} the (stochastic) k-armed bandit describes a sequential decision making problem, with $k$ possible actions or arms. Each arm $i$ is associated with a fixed but unknown reward distribution. \footnote{In order to quantify the performance of bandit algorithms, some assumptions are required on the distributions from which the rewards are generated. It sufficient (but not nessessary) to assume they are sub-Gaussian.} For each time step up to some horizon $T$ the learner selects an action and receives a reward, sampled i.i.d from the marginal distribution corresponding to that action. The goal of the learner is to maximise the total reward they receive. This problem captures the exploration-exploitation trade-off, the learner must balance playing arms that have yielded good results previously with exploring arms about which they are uncertain.

\vspace{0.5cm}
\begin{definition}[Stochasitic k-armed bandit problem]
Let $\actionspace = \set{1,...k}$ be the set of available actions (or bandit arms) and $\P{\vec{Y}} = \P{Y_1,...,Y_k}$ be the joint distribution over the rewards for each action. The multi-armed bandit problem proceeds over $T$ rounds. In each round $t$, 
\begin{enumerate}
\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
\item the world stochastically generates the rewards for each action, $[Y_{t,1},...,Y_{t,k}] \sim \P{\vec{Y}}$
\item the learner observes and recieves (only) the reward for the selected action $Y_{t,a_t}$ 
\end{enumerate}
At the end of the game the total reward obtained by the learner is $\sum_{t=1}^T Y_{t,a_t}$. We denote the expected reward for the action $i$, $\E{ \P{Y_i}}$, by $\mu_i$ and the action with the highest expected reward by $i^*$. 
\end{definition}

The total reward a bandit algorithm/policy can expect to acheive depends on the distributions from which the rewards for each action are sampled. To account for this, the performance of bandit algorithms is quantified by the the difference between the reward obtained by the algorithm and the reward that would have been obtained by an oracle that selects the arm with the highest expected reward at every time step. This difference is known as the (cummulative) regret \footnote{The term regret is somewhat overloaded in the reinforcement learning liturature. There are alternative definitions that arise in the related problems of adversarial bandits and learning from expert advice. In addition, researches often refer to the expected regret as \quotes{the regret}.},
\eqn{
\label{eqn:raw_regret}
\rawregret = \sum_{t=1}^T Y_{t,i^*} - \sum_{t=1}^T Y_{t,a_t}
}

Both the rewards and the actions selected by the algorithm are random variables. The majority of work in the bandit liturature focuses on analysing and optimising some form of the expected regret, however there has been some work that also considers the concentration of the regret \citep{Auer1995,Audibert2007,Audibert2009exploration}. The expectation of the regret, as defined by equation \ref{eqn:raw_regret}, is refered to as the pseudo-regret \citep{Bubeck2012} and is given by equation \ref{eqn:pseudo_regret}. A stochastic bandit algorithm is learning if it obtains psuedo-regret that is sub-linear in $T$.

@@

The regret is invariant to adding a constant to the expected rewards for all actions. However, it still depends on key characteristics of the reward distributions for each action. Bandit algorithms are designed given assumptions about the form of the distributions, such as that they come from a given family (i.e bernoulli bandits, gaussian bandits) or that the rewards are bounded in some range. Given these assumptions, the performance of the algorithm is characterised in two ways; by the \emph{problem dependent regret}, which typically depends on how far each arm is from optimal and by the \emph{worst case regret}, which is the maximum regret over all possible configurations of the reward distributions (for a given horizon $T$ and number of arms $k$).

\todo{consider mentioning Thompson sampling and Bayesian bandits}

@@

The adaptive nature of multi-armed bandit algorithms complicates the design and analyse of estimators. The action selected by an algorithm at a given timestep can depend on the history of previous actions and rewards. As a result, the probability that each action is selected evolves over time, the actions are not sampled i.i.d from a fixed distribution and the number of times each action is selected is a random variable. The expectation and variance guarentees of standard estimators do not hold in this setting, see figure \ref{fig:early_stopping} for a concrete example. This makes it very difficult to  obtain an analytical expression for the expected regret for a given algorithm and problem. Instead, the focus is on computing bounds on the expected regret.

\begin{figure}[h]
\centering
\caption{Standard empirical estimators can be biased if the number of samples, $n$, is not fixed in advance, but is a random variable that depends on the values of previous samples. This example plots the distribution (over $10^6$ simulations) of $\hat{\mu} = \frac{1}{n}\sum_{i=1}^n X_i$, where $X_i \sim bernoulli(0.2)$. In each simulation, we stop taking samples if the average value of $X_i$ up to that point exceeds a threshold of $0.3$ or $n$ reaches $100$. $\E{\hat{\mu}} = 0.439$. The estimator is substantially biased above $\E{X_i} = 0.2$ by the early stopping. Note that excluding experiments that were stopped early creates a bias in the opposite direction, $\E{\hat{\mu}|n=100} = 0.185$, as trials that obtained positive results early are excluded. This has some interesting potential real world implications. Early stopping of clinical trials is controversial. A researcher conducting a meta-analysis who wished to avoid (rather than bound) bias due to early stopping would have to exclude not only those trials which were stopped early but those which \emph{could} have been stopped early.}
\label{fig:early_stopping}
\includegraphics[scale=1]{extreme_early_stopping}
\end{figure}

There are a few key hueristic principles that are used to guide the development of bandit algorithms. The first is \emph{optimism in the face of uncertainty}. Applied to stochastic bandits, the optimism in the face of uncertainty princple suggests computing a plausible upper-bound for the expected reward of each arm, and selecting the arm with the highest upper bound. The optimism principle encourages exploitation and exploration because a high upper bound on the expected reward for an action implies either the expected reward or the uncertainty about the reward for that action is high. Thus selecting it yields either a good reward or useful information.

\citet{Lai1985} leveraged the optimism in the face of uncertainty principle to develop an algorithm for specific families of reward distributions, including the expontial family. They showed that, for a given bandit problem, the pseudo-regret increased with $\bigo{log(T)}$ asymptopically and proved this is asymptopically efficient. However, their algorithm is complex and memory intensive to compute as, at each timestep, it relies on the entire sequence of rewards for each arm. \citet{Agrawal1995} developed a simpler algorithm that computed upper bounds based only on the mean of previous samples for each arm, whist retaining the logarithmic dependence on $T$. Finally, \citet{Auer2002} developed the UCB-1 algorithm, see algorithm \ref{alg:ucb-1}, which requires only that the reward distributions are bounded, and proved finite-time regret bounds. We now assume the rewards are bounded in $[0,1]$. The algorithm and regret bounds can be generalised to subgaussian reward distributions, see \citet{Bubeck2012}.

@@

Let $\Delta_i = \mu_i - \mu^*$ be degree to which each arm is sub-optimal. The problem dependent psuedo-regret for UCB-1 is bounded by equation \ref{eqn:ucb_problem_dependent_regret} \citep{Bubeck2012},

@@

Somewhat unintuitively, the regret increases as the value of the arms gets closer together. This is because it becomes harder for the algorithm to identify the optimal arm. As the differences $\Delta_i \rightarrow 0$, the regret bound in \ref{eqn:ucb_problem_dependent_regret} blows up, however the regret itself does not - since although we may not be able to distinguish arms with very small $\Delta_i$ from the optimal arm, we also do not lose much by selecting them. The worst case occurs if all arms have the same expected reward $\mu$ except for the optimal arm which has reward $\mu^*=\mu+\Delta$, where $\Delta$ is just too small for the algorithm to learn to identify which arm is optimal given the horizon $T$. The regret cannot exceed what would be obtained by selecting the a sub-optimal arm in every timestep, $T\Delta$, so the worst case regret is bounded by the minimum of equation \ref{eqn:ucb_problem_dependent_regret} and $T\Delta$ which ia maximised when they are equal, see figure \ref{fig:ucb_regret_vs_delta}. By solving this equality for $\Delta$ one can show the worst case regret is bounded by equation \ref{eqn:ucb_worst_case_regret}, see \citet{Bubeck2012}.

@@

\begin{figure}
\caption{The regret bound in equation \ref{eqn:ucb_problem_dependent_regret} grows as the differences between the expected rewards for each arm shrink. The solid curve shows the mean (cummulative) regret, over a 1000 simulations for a 2-armed, bernoulli bandit with fixed horizon, $T = 500$, as a function of the difference in the expected reward for the arms $\Delta$. The dashed curves show the corresponding upper bounds; $T\Delta$ and equation \ref{eqn:ucb_problem_dependent_regret}}
\label{fig:ucb_regret_vs_delta}
\includegraphics[scale=1]{ucb_regret_vs_delta}
\end{figure}

@@

Subtle modifications to the UCB algorithm can eliminate the logarithmic term equation \ref{eqn:ucb_worst_case_regret}. This yeilds regret $ \bigo{\sqrt{TK}}$ and closes the gap with the worst case lower bound \cite{Audibert2009,Lattimore2015}, whilst retaining a good problem dependent bound of the form achieved by UCB \cite{Lattimore2015}.

@@

@@

\todo{discuss the key ways in which the simple regret problem differs from cummulative regret}

@@

Adversarial bandits are an alternate, widely studied, setting that relaxes the assumption that rewards are generated stochastically. Instead, simultaneously with the learner selecting an action $a_t$, a potentially malicious adversary selects the reward vector $\boldsymbol{Y}_t$. As in the stochastic setting, the learner then receives reward only for the selected action.

\vspace{0.5cm}
\begin{definition}[Adversarial k-armed bandit problem]
Let $\actionspace = \set{1,...k}$ be the set of available actions.  In each round $t \in 1,...,T$, 
\begin{enumerate}
\item the world (or adversary) generates, but does not reveal, a vector or rewards $\vec{Y_t}=[Y_{t,1},...,Y_{t,k}]$. 
\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
\item the learner observes and recieves (only) the reward for the selected action $Y_{t,a_t}$ 
\end{enumerate} 
\end{definition}

>>>>

\chapter*{To-do}
\begin{enumerate}
\item Add something about Granger causality - does this belong in chapter 2 or chapter 3?
\item Sell my PhD as two key contribution. 1) A technical one, the other educational and unifying (across viewpoints)
\item Add a chapter in 'PhD prose' making some statement about the intersection of prediction and causality. When is a problem causal, what bounds does predictive accuracy give on causal models? To what degree is predictive accuracy a metric on which causal models can be judged? 
\item Benchmark estimation algorithms
\item clarify causal interpretation of Dirichlet models
\item Empirical evaluation of bandit algorithms (updated to include latest suggestions). Maybe a proper software bench-marking solution?
\item If comparing bandit performance, it would be interesting to include some that violate the stochastic bandit assumption. For example, if the rewards change slowly over time or jump suddenly a limited number of times, or with some small probability. 
\end{enumerate}

@@

This implies the counterfactual distributions are equivalent to the corresponding conditional distributions and, for a binary outcome $Y$, the causal effect is,

@@

The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignorability assumptions \citep{Rosenbaum1983}. They state that the treatment each person receives is independent of whether they would recover if treated and if they would recover if not treated. This is justified in example \ref{exm:ranomized_experiment} due to the randomisation of treatment assignment. In general the treatment assignment will not be independent of the potential outcomes. In example \ref{exm:adjusting}, the children who attended preschool may be more likely to have graduated high school had they in fact not attended than the children who actually did not attend, $X \nci \cf{Y}{0}$. Similarly, had the poorer children who did not attend pre-school attended they might not have done as well as the children who did in fact attend, $X \nci \cf{Y}{1}$. A more general form of the ignorability assumption is to identify a set of variables $Z$ such that $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$.

@@

\begin{figure}[h]
\centering
\caption{
Parametric regression may yield poor estimates of causal effects if the model is miss-specified, even if the model fits well over the domain of the training data. In this example, $\P{Z|X=0} \sim N(\mu_0,\sigma_0^2)$ and $\P{Z|X=1} \sim N(\mu_1,\sigma_1^2)$ with little overlap in the densities. If $X=0$ then $Y \sim N(f_1(x)=sin(x),\sigma_y^2)$ and if $X=1$ then $Y \sim N(f_0(x) = \frac{1}{x+1},\sigma_y^2)$. We estimate $f_1(z)$ from the sample in which $X=1$ (green points) and $f_0(z)$ from the sample for which $X=0$ (blue points). In both cases the linear model is a good fit to the data. However, the resulting estimate of the causal effect is very poor for the lower values of $z$. } 
\label{fig:covariate_shift_adjust}
\includegraphics[scale=.55]{covariate_shift_estimation.pdf}
\end{figure}

@@

The propensity score can be used to estimate the average causal effect without specifying a regression model for $\E{Y|X,\vb{Z}}$. \citet{Rosenbaum1983} demonstrated that if the ignorability assumption is satisfied by conditioning on $\vb{Z}$, then it is also satisfied by conditioning on $e(\vb{z})$. This allows for estimators based on stratifying, matching or regression on the propensity score rather than the covariates $\vb{Z}$. Inverse propensity weighting can also be combined with empirical estimation of $\E{Y|X,Z}$ yielding the simple, albeit inefficient, estimator in equation \ref{eqn:simple_inverse_propensity} \citep{Imbens2004}. In some settings, such as stratified randomised trials \citep{imbens2015causal} or learning from logged bandit feedback \citep{Bottou2013} the propensity score may be known. However in general, it must be estimated from data. Frequently this is done with a simple parametric model such as logistic regression, but a wide range of standard machine learning algorithms including bagging and boosting, random forests and neural networks can also be applied   \citep{Austin2011}. \citet{Lunceford2004} review the theoretical properties of key propensity score based estimators, including stratification and inverse propensity weighting.

@@

The performance of methods for causal effect estimation can be tested on simulated data \citep{Frolich2001,zhao2004using,Hill2011,Dorie2016} or by comparing estimates from observational studies with the results from corresponding experiments \citep{lalonde1986evaluating,Fraker1987,heckman1997matching,heckman1998characterizing,dehejia1999causal,Smith2001,Anglemyer2014}. Unfortunately there are a relatively small number of examples where comparable observational and experimental data are available. The results are mixed with later studies finding generally better alignment of results but it is hard to ascertain if this is due to improved methodological approaches or over-fitting to the available data-sets.

@@

The previous sections all focus on aspects of the question; how can we estimate the effect of an intervention in a system from data collected prior to taking it. There is an obvious alternative. Instead of trying to infer the outcome of an intervention from passive observations one can intervene and see what happens. There are three key differences between observing a system and explicitly intervening in it. Firstly, when we intervene we choose which actions to take and thus have control over which data points we obtain to learn from. Selecting data points optimally for learning is the focus of the optimal experimental design literature within statistics \citep{pukelsheim2006optimal} and the active learning literature in machine learning \citep{settles2010active}. Secondly, explicitly choosing interventions yields a perfect model of the probability with which each action is selected, given any context, allowing control over confounding bias. Finally, when we are intervening in a system we typically care about the impact of our actions on the system in addition to optimising learning. For example, in a drug trial , assigning people a sub-optimal treatment has real world costs. This leads to a trade-off between exploiting the best known action so far and exploring alternative actions about which we are less certain. This exploration-exploitation trade-off lies at the heart of the field of reinforcement learning \citep{sutton1998reinforcement}.

@@

Stratified randomised experiments address the issue of variance due to covariate imbalance by randomly allocating treatment conditional on covariates believed to influence the outcome of interest. This introduces a backdoor path from treatment to outcome via the covariates on which treatment is stratified, see figure \ref{fig:random_experiment_network_stratified}, necessitating that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under under ignorability \S\ref{sec:causal_effect_under_ignorability}. The key difference is that the propensity score is known, as it is designed by the experimenter, and there are guaranteed (rather than assumed) to be no latent confounding variables (that influence both treatment and outcome). See \citet{imbens2015causal} for a discussion of the trade-offs between stratified versus completely random experiments.

IF THE PROPENSITY SCORE IS THE SAME FOR ALL COVARIATES IT DOESN'T SEEM ONE MUST ADJUST.
\todo{double check}

@@

It is not always possible or ethical to conduct a randomised controlled trial as is beautifully demonstrated by the paper of \citet{smith2003parachute} on randomised cross-over trials of parachute use for the reduction of the mortality and morbidity associated with falls from large heights \ref{fig:parashute}. When experimentation is possible, it is frequently difficult or expensive. This means experimental data sets are often much smaller than observational ones, limiting the complexity of models we can explore. In addition, they are often conducted on a convenient but unrepresentative sample of the broader population of interest (for example first year students at research universities). This can result in estimates with high \emph{internal validity} \citep{Campbell1963} in that they should replicate well in a similar population, but very low \emph{external validity}; the results may not carry over to the general population of interest. The question of whether an experiment conducted on one population can be mapped to another is referred to as the transportability problem \citep{Bareinboim2013} and relies on very similar assumptions and arguments to causal inference and the do-calculus.

\begin{figure}
\begin{center}
\caption{Experiments are not always ethical; an illustration of a randomised cross-over trial of parachutes for the prevention of morbidity and mortality associated with falls from large heights.}
\label{fig:parashute}
\includegraphics[scale=.4]{double_blind_parashute_trial_cartoon.png}
\end{center}
\end{figure}

@@

Multi-armed bandits address the problem of designing experiments that can adapt as samples are observed. Their introduction is generally attributed to \citet{Thompson1993}. In its classic formulation \cite{Robbins1952,Lai1985} the (stochastic) k-armed bandit describes a sequential decision making problem, with $k$ possible actions or arms. Each arm $i$ is associated with a fixed but unknown reward distribution. \footnote{In order to quantify the performance of bandit algorithms, some assumptions are required on the distributions from which the rewards are generated. It sufficient (but not necessary) to assume they are sub-Gaussian.} For each time step up to some horizon $T$ the learner selects an action and receives a reward, sampled i.i.d from the marginal distribution corresponding to that action. The goal of the learner is to maximise the total reward they receive. This problem captures the exploration-exploitation trade-off, the learner must balance playing arms that have yielded good results previously with exploring arms about which they are uncertain.

\vspace{0.5cm}
\begin{definition}[Stochasitic k-armed bandit problem]
Let $\actionspace = \set{1,...k}$ be the set of available actions (or bandit arms) and $\P{\vec{Y}} = \P{Y_1,...,Y_k}$ be the joint distribution over the rewards for each action. The multi-armed bandit problem proceeds over $T$ rounds. In each round $t$, 
\begin{enumerate}
\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
\item the world stochastically generates the rewards for each action, $[Y_{t,1},...,Y_{t,k}] \sim \P{\vec{Y}}$
\item the learner observes and receives (only) the reward for the selected action $Y_{t,a_t}$ 
\end{enumerate}
At the end of the game the total reward obtained by the learner is $\sum_{t=1}^T Y_{t,a_t}$. We denote the expected reward for the action $i$, $\E{ \P{Y_i}}$, by $\mu_i$ and the action with the highest expected reward by $i^*$. 
\end{definition}

The total reward a bandit algorithm/policy can expect to achieve depends on the distributions from which the rewards for each action are sampled. To account for this, the performance of bandit algorithms is quantified by the the difference between the reward obtained by the algorithm and the reward that would have been obtained by an oracle that selects the arm with the highest expected reward at every time step. This difference is known as the (cumulative) regret \footnote{The term regret is somewhat overloaded in the reinforcement learning literature. There are alternative definitions that arise in the related problems of adversarial bandits and learning from expert advice. In addition, researches often refer to the expected regret as \quotes{the regret}.}.
\eqn{
\label{eqn:raw_regret}
\rawregret = \sum_{t=1}^T Y_{t,i^*} - \sum_{t=1}^T Y_{t,a_t}
}

Both the rewards and the actions selected by the algorithm are random variables. The majority of work in the bandit literature focuses on analysing and optimising some form of the expected regret, however there has been some work that also considers the concentration of the regret \citep{Auer1995,Audibert2007,Audibert2009exploration}. The expectation of the regret, as defined by equation \ref{eqn:raw_regret}, is referred to as the pseudo-regret \citep{Bubeck2012} and is given by equation \ref{eqn:pseudo_regret}. A stochastic bandit algorithm is learning if it obtains pseudo-regret that is sub-linear in $T$.

@@

The regret is invariant to adding a constant to the expected rewards for all actions. However, it still depends on key characteristics of the reward distributions for each action. Bandit algorithms are designed given assumptions about the form of the distributions, such as that they come from a given family (i.e Bernoulli bandits, Gaussian bandits) or that the rewards are bounded in some range. Given these assumptions, the performance of the algorithm is characterised in two ways; by the \emph{problem dependent regret}, which typically depends on how far each arm is from optimal and by the \emph{worst case regret}, which is the maximum regret over all possible configurations of the reward distributions (for a given horizon $T$ and number of arms $k$).

@@

The adaptive nature of multi-armed bandit algorithms complicates the design and analyse of estimators. The action selected by an algorithm at a given time-step can depend on the history of previous actions and rewards. As a result, the probability that each action is selected evolves over time, the actions are not sampled i.i.d from a fixed distribution and the number of times each action is selected is a random variable. The expectation and variance guarantees of standard estimators do not hold in this setting, see figure \ref{fig:early_stopping} for a concrete example. This makes it very difficult to  obtain an analytical expression for the expected regret for a given algorithm and problem. Instead, the focus is on computing bounds on the expected regret.

\begin{figure}[h]
\centering
\caption{Standard empirical estimators can be biased if the number of samples, $n$, is not fixed in advance, but is a random variable that depends on the values of previous samples. This example plots the distribution (over $10^6$ simulations) of $\hat{\mu} = \frac{1}{n}\sum_{i=1}^n X_i$, where $X_i \sim Bernoulli(0.2)$. In each simulation, we stop taking samples if the average value of $X_i$ up to that point exceeds a threshold of $0.3$ or $n$ reaches $100$. $\E{\hat{\mu}} = 0.439$. The estimator is substantially biased above $\E{X_i} = 0.2$ by the early stopping. Note that excluding experiments that were stopped early creates a bias in the opposite direction, $\E{\hat{\mu}|n=100} = 0.185$, as trials that obtained positive results early are excluded. This has some interesting potential real world implications. Early stopping of clinical trials is controversial. A researcher conducting a meta-analysis who wished to avoid (rather than bound) bias due to early stopping would have to exclude not only those trials which were stopped early but those which \emph{could} have been stopped early.}
\label{fig:early_stopping}
\includegraphics[scale=1]{extreme_early_stopping}
\end{figure}

There are a few key principles that are used to guide the development of bandit algorithms. The simplest is to explicitly separate exploration from exploitation and base estimation of the expected rewards of each arm only on the data generated during exploration steps. A common example in practice is uniform exploration (or A/B testing) for some fixed period followed by selecting the action found to be best during the exploration phase. This results in simpler analysis, particularly if the number of exploration steps is fixed in advance, however it is sub-optimal, even if the exploration period is adaptive \citep{NIPS2016_6179}.

Another key approach is \emph{optimism in the face of uncertainty}. Applied to stochastic bandits, the optimism in the face of uncertainty principle suggests computing a plausible upper-bound for the expected reward of each arm, and selecting the arm with the highest upper bound. The optimism principle encourages exploitation and exploration because a high upper bound on the expected reward for an action implies either the expected reward or the uncertainty about the reward for that action is high. Thus selecting it yields either a good reward or useful information.

\citet{Lai1985} leveraged the optimism in the face of uncertainty principle to develop an algorithm for specific families of reward distributions, including the exponential family. They showed that, for a given bandit problem, the pseudo-regret increased with $\bigo{log(T)}$ asymptotically and proved this is asymptotically efficient. However, their algorithm is complex and memory intensive to compute as, at each time-step, it relies on the entire sequence of rewards for each arm. \citet{Agrawal1995} developed a simpler algorithm that computed upper bounds based only on the mean of previous samples for each arm, whist retaining the logarithmic dependence on $T$. Finally, \citet{Auer2002} developed the UCB-1 algorithm, see algorithm \ref{alg:ucb-1}, which requires only that the reward distributions are bounded, and proved finite-time regret bounds. We now assume the rewards are bounded in $[0,1]$. The algorithm and regret bounds can be generalised to subgaussian reward distributions, see \citet{Bubeck2012}.

@@

Let $\Delta_i = \mu_i - \mu^*$ be degree to which each arm is sub-optimal. The problem dependent pseudo-regret for UCB-1 is bounded by equation \ref{eqn:ucb_problem_dependent_regret} \citep{Bubeck2012},

@@

Somewhat unintuitively, the regret increases as the value of the arms gets closer together. This is because it becomes harder for the algorithm to identify the optimal arm. As the differences $\Delta_i \rightarrow 0$, the regret bound in \ref{eqn:ucb_problem_dependent_regret} blows up, however the regret itself does not - since although we may not be able to distinguish arms with very small $\Delta_i$ from the optimal arm, we also do not lose much by selecting them. The worst case occurs if all arms have the same expected reward $\mu$ except for the optimal arm which has reward $\mu^*=\mu+\Delta$, where $\Delta$ is just too small for the algorithm to learn to identify which arm is optimal given the horizon $T$. The regret cannot exceed what would be obtained by selecting the a sub-optimal arm in every time-step, $T\Delta$, so the worst case regret is bounded by the minimum of equation \ref{eqn:ucb_problem_dependent_regret} and $T\Delta$ which ia maximised when they are equal, see figure \ref{fig:ucb_regret_vs_delta}. By solving this equality for $\Delta$ one can show the worst case regret is bounded by equation \ref{eqn:ucb_worst_case_regret}, see \citet{Bubeck2012}.

@@

\begin{figure}
\caption{The regret bound in equation \ref{eqn:ucb_problem_dependent_regret} grows as the differences between the expected rewards for each arm shrink. The solid curve shows the mean (cumulative) regret, over a 1000 simulations for a 2-armed, Bernoulli bandit with fixed horizon, $T = 500$, as a function of the difference in the expected reward for the arms $\Delta$. The dashed curves show the corresponding upper bounds; $T\Delta$ and equation \ref{eqn:ucb_problem_dependent_regret}}
\label{fig:ucb_regret_vs_delta}
\includegraphics[scale=1]{ucb_regret_vs_delta}
\end{figure}

@@

Subtle modifications to the UCB algorithm can eliminate the logarithmic term equation \ref{eqn:ucb_worst_case_regret}. This yields regret $ \bigo{\sqrt{TK}}$ and closes the gap with the worst case lower bound \cite{Audibert2009,Lattimore2015}, whilst retaining a good problem dependent bound of the form achieved by UCB \cite{Lattimore2015}.

@@

Finally, there is the heuristic principle of playing each arm with probability proportional to the likelyhood that it is optimal. This approach is generally called Thompson sampling as it was the method proposed in the original bandit paper by \citet{Thompson1993}. Thompson sampling has strong empirical performance, \citep{chapelle2011empirical}. However, it is complex to analyse, \citet{kaufmann2012thompson} demonstrate that it obtains optimal problem dependent bounds, \citet{agrawal2013further} show that it obtains worst case regret of $\bigo{\sqrt{kT\log(T)}}$, equivalent to UCB.

@@

\todo{discuss the key ways in which the simple regret problem differs from cumulative regret}

@@

Adversarial bandits, described by \citet{Auer1995}, are an alternate, widely studied, setting that relaxes the assumption that rewards are generated stochastically. Instead, simultaneously with the learner selecting an action $a_t$, a potentially malicious adversary selects the reward vector $\boldsymbol{Y}_t$. As in the stochastic setting, the learner then receives reward only for the selected action.

\vspace{0.5cm}
\begin{definition}[Adversarial k-armed bandit problem]
\label{def:adversarial-k-armed-bandit}
Let $\actionspace = \set{1,...k}$ be the set of available actions.  In each round $t \in 1,...,T$, 
\begin{enumerate}
\item the world (or adversary) generates, but does not reveal, a vector or rewards $\vec{Y_t}=[Y_{t,1},...,Y_{t,k}]$. 
\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
\item the learner observes and receives (only) the reward for the selected action $Y_{t,a_t}$ 
\end{enumerate} 
\end{definition}

Adversaries that generate rewards independently of the sequence of actions selected by the learner in previous time steps are referred to as \emph{oblivious}, as opposed to \emph{non-oblivious} adversaries, which can generate rewards as a function of the history of the game. In the case of oblivious adversaries, we can also define the adversarial bandit problem by assuming the adversary generates the entire sequence of reward vectors before the game commences.
